---
title: "Gamm4Tutorial"
author: "Michael Andreae"
date: "February 22, 2016"
output: pdf_document
---

# Introduction second chapter
This is the second chapter of our tutorial on analysing the cognitive decline dataset by Drs. Lipton and Hall using a generalized additive mixed models inspired and guided by the book \href{http://www.highstat.com/BGGAMM.htm}{A Beginner's Guide to Generalized Additive Mixed Models with R. Zuur AF, Saveliev AA, Ieno EN.}

## Setting up
```{r knitr, echo=FALSE}
suppressWarnings(library(knitr))
```

```{r globle_knitr_options, echo=FALSE}
rm(list=ls())
suppressWarnings(library(knitr))
# set options
opts_chunk$set(fig.width=6, fig.height=4, 
               fig.align = "left",
               fig.path='Figs/',
               echo=FALSE, warning=FALSE, message=FALSE)
eval_everything=FALSE
```

```{r library}

library(ggplot2)
library(mgcv)
library(lattice)
# library(parallel)
# require(knitr)
# library(tidyr)
# library(rstan)
# library(rstanarm)
# library(shinystan)
# library(nlme)
# library(dplyr)
```

## Sourcing support files
We source the support files available \href{http://www.highstat.com/BGGAMM.htm#Data_sets_and_R_code_used}{online}

```{r source_support_files}
source(file = "HighstatLibV6.R") 
source(file = "MCMCSupportHighstatV2.R")
source(file = "SupportFilesHighstat.R")
```

# Import and explore Data
## Import and load squid data
We import \href{http://www.highstat.com/BGS/GAMM/Data/SquidNorway.txt}{the polar bear dataset} from the website and save it as *PolarBearsV2.Rdata* after we factorize certain variables

```{r import_polar_bear_data, eval=FALSE}
PB <- read.table(file= "PolarBearsV2.txt", header = TRUE, dec = ".")

PB$fBearID <-factor(PB$BearID)
PB$fRepro <-factor(PB$Repro)
PB$fDen <-factor(PB$Den)
PB$fMonth <-factor(PB$Month)
PB$fSeason <-factor(PB$Season)

save(PB, file="PolarBearsV2.Rdata")
```

## Load polar bear data
We load the *PolarBearsV2.Rdata* file.
```{r load_polarbear}
load("PolarBearsV2.Rdata")
str(PB)
```

# Exploration
## Cleveland dotplot
```{r Cleveland_plot}
c <- ggplot(PB, aes(x=Movement, y=1:nrow(PB)))
c +geom_point()
```

## Changes in movement over time
```{r changes_movement_time}
ch <- ggplot(PB, aes(x=dDay, y=Movement))
ch +geom_point() +stat_smooth() +theme_minimal()
```

## Boxplot for seasonality
Investigating the amount of movement in different months.
```{r boxplot_season}
b <- ggplot(PB, aes(fMonth, Movement))
b + geom_boxplot()
```

# Model
## Centering
```{r centering_polarbear}
PB$Yearc <- PB$Year-mean(PB$Year)
PB$DayInYearc <- PB$DayInYear - mean(PB$DayInYear)
```

## Frequentist approach

### Model
```{r frequentist_model}
M1 <- gamm(Movement ~ fRepro +Yearc +s(DayInYearc), data=PB,
           random = list(BearID=~ 1))
```

### Results
```{r frequentist_results}
summary(M1$gam)
anova(M1$gam)
summary(M1$lme)
```

### Smoother plot
```{r frequentist_smoother}
plot(M1$gam, cex.lab = 1.5, axes=FALSE, xlab="Day of the year")
```