---
title: "Gamm4Tutorial"
author: "Michael Andreae"
date: "February 22, 2016"
output: pdf_document
---

# Introduction
This is a tutorial on analysing the cognitive decline dataset by Drs. Lipton and Hall using a generalized additive mixed models inspired and guided by the book \href{http://www.highstat.com/BGGAMM.htm}{A Beginner's Guide to Generalized Additive Mixed Models with R. Zuur AF, Saveliev AA, Ieno EN.}

## Setting up
```{r knitr, echo=FALSE}
suppressWarnings(library(knitr))
```

```{r globle_knitr_options, echo=FALSE}
rm(list=ls())
suppressWarnings(library(knitr))
# set options
opts_chunk$set(fig.width=6, fig.height=4, 
               fig.align = "left",
               fig.path='Figs/',
               echo=FALSE, warning=FALSE, message=FALSE)
eval_everything=FALSE
```

```{r library}

library(ggplot2)
library(mgcv)
library(lattice)
# library(parallel)
# require(knitr)
# library(tidyr)
# library(rstan)
# library(rstanarm)
# library(shinystan)
# library(nlme)
# library(dplyr)
```

## Sourcing support files
We source the support files available \href{http://www.highstat.com/BGGAMM.htm#Data_sets_and_R_code_used}{online}

```{r source_support_files}
source(file = "HighstatLibV6.R") 
source(file = "MCMCSupportHighstatV2.R")
source(file = "SupportFilesHighstat.R")
```

# Import and explore Data
## Import and load squid data
We import \href{http://www.highstat.com/BGS/GAMM/Data/SquidNorway.txt}{the squid dataset} from the website and save it as *SquidNorway.Rdata*

```{r import_squid_data, eval=FALSE}
Squid <- read.table(file= "SquidNorway.txt", header = TRUE, dec = ".")
save(Squid, file="SquidNorway.Rdata")
```

## Load SquidNorway data
We load the *SquidNorway.Rdata* file.
```{r load_SquidNorway}
load("SquidNorway.Rdata")
str(Squid)
```

## Explore squid data
We plot the outcome *d15N* radioisotope versus the predictors.
```{r exploreSquiddata}
Y    <- Squid$d15N
MyX  <- c("Lat", "Depth", "ML")
X    <- Squid[, MyX]

AllX  <- as.vector(as.matrix(Squid[,MyX]))
AllY  <- rep(Squid[,"d15N"] , length(MyX))
AllID <- rep(MyX, each = nrow(Squid))
    
xyplot(AllY ~ AllX|factor(AllID), col = 1,layout=c(3,1),
              xlab = list(label="Explanatory variables",cex = 1.5),
              ylab = list(label = "d15N", cex = 1.5),
              strip = function(bg='white', ...)
                strip.default(bg='white', ...),
              scales = list(alternating = T,
                            x = list(relation = "free"),
                            y = list(relation = "same")),
              panel=function(x, y){
                panel.grid(h=-1, v= 2)
                panel.points(x, y, col = 1)
                panel.loess(x, y, col = 1, lwd = 2)})
```

# Modeling
## Linear Model
We fit a linear regression, plot the standardized residuals against the predictor length *ML*
```{r linearSquidModel}
M1 <- lm(d15N ~ Lat +ML, data=Squid)
Squid$E1 <- rstandard(M1)

p <- ggplot(Squid, aes(ML, E1)) +
  geom_point() +
  geom_smooth()
p +xlab("ML") +
  ylab("standardized residuals") +theme_minimal() +
  geom_hline(yintercept = 0)
```

## Fit GAM
We fit a first generalized additive model using *gam* from the package *mgcv*, using different smoothers and plot the smoothers
```{r first_GAM}
M2A <- gam(d15N ~ Lat +s(ML), data=Squid)
M3A <- gam(d15N ~ Lat +s(ML, bs="cr"), data=Squid)
M4A <- gam(d15N ~ Lat +s(ML, bs ="ps"), data=Squid)
M5A <- gam(d15N ~ Lat +s(ML, bs="cs"), data=Squid)

plot(M2A)
plot(M3A)
plot(M4A)
plot(M5A)
```

## Standardize covariates
We standardize the covariates to improve mixing for MCMC.
```{r standardize_covariates}
# need to source the routines source files included with the book
Squid$Lat.std<-Mystd(Squid$Lat)
Squid$ML.std<-Mystd(Squid$ML)
```

## Second GAM
We fit a second gam with the standardized residuals and plot the smoother.
```{r second_GAM}
M3 <- gam(d15N ~ Lat.std + s(ML.std, bs="cr"), data=Squid)
plot(M3, residuals = TRUE, , pch=16, cex=0.7)
summary(M3)
```

# Linear spline regression
## Linear spline function
```{r linear_spline_function, echo=TRUE}
rhs <- function(x, TH) ifelse(x >= TH, x-TH, 0)
```

## Linear spline model
We fit a model with a linear spline with one knot.
```{r linear_spline_model}
M4 <- gam(d15N ~ Lat.std +ML.std +rhs(ML.std, 0.8), data=Squid)
summary(M4)
```

## Ploting the linear spline
We visualize the predicted line by predicting on a new dataset holding $Lat.std = 0$ and plotting this as a red line.
```{r plotting_linear_spline_singleKnot}
NewData<- data.frame(ML.std = seq(from = -1.48, to = 3.03, 
                                  length = 100),
                     Lat.std=0)
P4 <- predict(M4, newdata = NewData)

## using build in plot
# plot(x=Squid$ML.std,y=Squid$d15N, pch=16, xlab="standardized ML",
#     ylab="d15N")
# abline(v=0.83, lty=2, lwd=2)
# lines(x=NewData$ML.std, y=P4, lwd=5)

## ggplot
l <- ggplot(Squid) +geom_point(aes(x=ML.std, y=d15N)) +
  geom_vline(xintercept = 0.83, 
               colour="green", linetype = "longdash")

l +geom_path(data = NewData, 
             aes(x=ML.std, y=P4),
              colour="red") +
  theme_minimal()
```

