---
title: "Gamm4Tutorial"
author: "Michael Andreae"
date: "February 22, 2016"
output: pdf_document
---

# Introduction second chapter
This is the second chapter of our tutorial on analysing the cognitive decline dataset by Drs. Lipton and Hall using a generalized additive mixed models inspired and guided by the book \href{http://www.highstat.com/BGGAMM.htm}{A Beginner's Guide to Generalized Additive Mixed Models with R. Zuur AF, Saveliev AA, Ieno EN.}

## Setting up
```{r knitr, echo=FALSE}
suppressWarnings(library(knitr))
```

```{r globle_knitr_options, echo=FALSE}
rm(list=ls())
suppressWarnings(library(knitr))
# set options
opts_chunk$set(fig.width=6, fig.height=4, 
               fig.align = "left",
               fig.path='Figs/',
               echo=FALSE, warning=FALSE, message=FALSE)
eval_everything=FALSE
```

```{r library}

library(ggplot2)
library(mgcv)
library(lattice)
# library(parallel)
# require(knitr)
# library(tidyr)
# library(rstan)
# library(rstanarm)
# library(shinystan)
# library(nlme)
# library(dplyr)
```

## Sourcing support files
We source the support files available \href{http://www.highstat.com/BGGAMM.htm#Data_sets_and_R_code_used}{online}

```{r source_support_files}
source(file = "HighstatLibV6.R") 
source(file = "MCMCSupportHighstatV2.R")
source(file = "SupportFilesHighstat.R")
```

# Import and explore Data
## Import and load squid data
We import \href{http://www.highstat.com/BGS/GAMM/Data/SquidNorway.txt}{the squid dataset} from the website and save it as *SquidNorway.Rdata*

```{r import_polar_bear_data, eval=FALSE}
PB <- read.table(file= "PolarBearsV2.txt", header = TRUE, dec = ".")
save(PB, file="PolarBearsV2.Rdata")
```

## Load polar bear data
We load the *PolarBearsV2.Rdata* file.
```{r load_polarbear}
load("PolarBearsV2.Rdata")
str(PB)
```
